---
title: "Supplementary Data Analysis - Random Forest Classification"
author: "Kristen N. Finch"
date: "12/1/2020"
output:
  html_document: default
  pdf_document: default
---

### About

This document was prepared by Kristen N. Finch to supplement the article. The reader will find R code necessary to replicate our random forest classification for bigleaf maple Level III Ecoregions using DNA genotypes with the files provided.

This dataset is available as supplement for a research article that was submitted in Decemebr 2020 to *Forensic Science International* special edition "Animals and Environments." 

Suggested citation for the Article: 
Cronn RC, Finch KN, Hauck LL, Parker-Forney M, Milligan B, Dowling J, and Adventure Scientists (2020). for Range-wide assessment of a SNP panel for individualization and geolocalization of bigleaf maple (*Acer macrophyllum* Pursh). Forensic Sci Int. Submitted.

This document and analysis was prepared by Kristen N. Finch. Questions about this document should be directed to Kristen N. Finch using the contact email provided under this ORCID ID: https://orcid.org/0000-0003-2098-7546 This document is not intended as a tutorial. This document is intended for intermediate and advanced users. 

All users should frequently examine generated data frames with head(). The View() function can also be used, but viewing large datasets can cause R to crash. 

### Set Up

This document was originally prepared with R version 3.6.1
(2019-07-05) -- "Action of the Toes" and R Studio version 1.2.1578. Analyses may not work as described with different versions of R and packages. *Also note that any changes to directory names or filenames will require changes to the code below.* I recommend using the filenames and directory names I have chosen to limit confusion.

To use the files provided, set your working directory to directory downloaded from *Forensic Science International* with setwd(PATH).

```{r,message=FALSE,warning=FALSE}
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
# ~/Cronn_etal_2020_supplement
# it will be necessary to change the path to the files. 
```

### Libaries

Dependencies may need to be installed manually. R will prompt. 

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(randomForest)
library(reshape2)
library(stringi)
library(stringr)
library(tidyr)
```

### Functions
This document contains functions that were developed to summarize data. 

####get_mode()
Used in the get_results finction below. Calculates mode for a column of data | Column of data frame.


```{r}
getmode<-function(column_of_data) {
  uniqv <- unique(column_of_data)
  uniqv[which.max(tabulate(match(column_of_data, uniqv)))]
}
```

#### get_results()

Calculates mean, median, mode, min, max, and 95% confidence interval for a column of data | Note on data & column of data frame.

Example Usage:

err_obs_result<-get_results("observed classification error",rf_results$err_obs)

Note_about_stat must follow character syntax with double quotes: "Note about stat"

```{r}
get_results<-function(note_about_stat,column_of_data){
  mean<-mean(column_of_data,na.rm = TRUE)
  median<-median(column_of_data,na.rm = TRUE)
  mode<-getmode(column_of_data)
  min<-min(column_of_data,na.rm = TRUE)
  max<-max(column_of_data,na.rm = TRUE)
  err<-qt(0.975,df=length(column_of_data)-1)*sd(column_of_data,na.rm = TRUE)/sqrt(length(column_of_data))
  up<-mean+err
  lo<-mean-err
  note_about_stat<-note_about_stat
  results<-cbind.data.frame(note_about_stat,mean,median,mode,min,max,err,up,lo)
  return(results)
}
```

### Data

Here we use the adegenet filtered data to remove individuals with greater than 10% missing data. 

```{r,message=FALSE,warning=FALSE,comment=""}
maple<-read.csv("~/Cronn_etal_2020_supplement/R_analysis/random_forest/MapleSNP_adegenet_1191i133s_filtered_1col.csv",header=1)
pop_tab<-data.frame(t(table(maple$pop))) #table establishes the minimum class size to maintain balanced class sizes for random forest analysis. 
maple<-maple[c(-1)] #sample ID's not needed for random forest classificiation.
```


### Regional Classification 

Run Random Forest Classification with Level 3 EcoRegions. *This analysis take 1.5 hours to complete.* 

Random forest require complete data. Before employing this analysis, I examined the effect of imputation method. 

Methods compared: 
1. imputation via randomForest parameter "na.action=na.roughfix" 
2. imputation method employed by R package synbreed (deprecated)

Comparison: 
Random forest classification 15 times with each imputation method. 

Conclusion: 
I found no difference in mean error comparing methods. I selected the random forest imputation method because it is the simpler method (less code), and is part of a package that is maintained. 

We estimated the range of classification errors for geographic localization by calculating the mean of the median error (%) across 5,000 random forests (2,500,000 total trees). The random expected error of a classification analysis is the reciprocal of the number of classes in the grouping variable; for example, classification analysis with 12 ecoregions should give a random classification error of (1 – (1/12)=) ~91.7%. For our analysis, we empirically estimated baseline error by randomizing the level 3 ecoregions, and comparing observed classification error to that from randomizations. 

```{r,eval=FALSE}

#oberved----
fit.list<-list()

#this will take about 1.5 hours to complete
for (i in 1:length(c(1:5000))){
  #set up loop
  cat("Random Forest Repetition number ") 
  cat(i)
  cat("\n")
  #set up training set such that class sizes are balanced and sample size is the number of samples in the smallest group.
  training_set<-maple%>%group_by(pop)%>%sample_n(min(pop_tab$Freq),replace=FALSE)
  fit.list[[i]]<-(fit.forest<-randomForest(pop~.,data=training_set,importance=TRUE,ntree=500, mtry=ncol(training_set)-1,na.action=na.roughfix))
}

#extract median error rates from the analysis
fit.median<-list()
for (i in 1:length(fit.list)){
  fit.median[[i]] <- median(fit.list[[i]]$err.rate[,'OOB'])
}

#extract class error rates from the analysis
class.errs<-list()
for (i in 1:length(fit.list)){
  class.errs[[i]] <- fit.list[[i]]$confusion[,'class.error']
}

#extract importance estimations from the analysis
gini.list<-list()
for (i in 1:length(fit.list)){
  gini.list[[i]] <- fit.list[[i]]$importance
}

#save results
err.obs<-data.frame(unlist(fit.median))
names(err.obs)[1]<-"error"
err.class<-melt(do.call("rbind",class.errs))
names(err.class)<-c("iter","pop","error")
ginis<-do.call("rbind",gini.list)

#OPTIONAL
#write results out to file 
#write.csv(err.obs,"yourfile_rf_res_obs.csv",quote=FALSE,row.names = FALSE)
#write.csv(err.class,"yourfile_rf_class_err.csv",quote=FALSE,row.names=FALSE)
#write.csv(ginis,"yourfile_rf_ginis.csv",quote=FALSE,row.names=TRUE)

#randomized----

#perform the same with randomized eco-regions (here "pop") to obtain an estimate for the baseline error rate.
rand.list<-list()

#this will take about 2 hours to complete
for (i in 1:length(c(1:5000))){
  #set up loop
  cat("Random Forest Repetition number ") 
  cat(i)
  cat("\n")
  #set up training set such that class sizes are balanced and sample size is the
  #number of samples in the smallest group.
  training_set<-maple%>%group_by(pop)%>%sample_n(min(pop_tab$Freq),replace=FALSE)
  rand.col<-data.frame(training_set$pop)
  training_set$pop<-NULL
  rand.col<-data.frame(sample_n(rand.col,nrow(rand.col),replace = FALSE))
  training_set<-cbind.data.frame(rand.col,training_set)
  names(training_set)[1]<-"pop"
  rand.list[[i]]<-(fit.forest<-randomForest(pop~.,data=training_set,importance=TRUE,ntree=500,
                                            mtry=ncol(training_set)-1,na.action=na.roughfix))
}

rand.median<-list()
for (i in 1:length(rand.list)){
  rand.median[[i]] <- median(rand.list[[i]]$err.rate[,'OOB'])
}

err.rand<-data.frame(unlist(rand.median))
names(err.rand)[1]<-"error"

#OPTIONAL
#write results out to file names of your choice
#write.csv(err.rand,"yourfile_rf_res_rand.csv",quote=FALSE,row.names=FALSE)

#OPTIONAL
#save the random forests so that you don't have to run them again to extract
#more values for visualization of results.
#save(fit.list, file="yourfile_ACMA_rf_obs_results.RData")
#save(rand.list, file="yourfile_ACMA_rf_rand_results.RData")
```

### Results

```{r,message=FALSE,warning=FALSE,comment=""}

#load data
err.obs<-read.csv("20200325_rf_res_obs.csv",header=1)
err.rand<-read.csv("20200325_rf_res_rand.csv",header=1)
err.class<-read.csv("20200325_rf_class_err.csv",header=1)

#add column to differentiate results generated with observed or randomized data
err.obs$group<-"obs"
err.rand$group<-"rand"

names(err.obs)
names(err.rand)

#convert to percent
err.obs<-mutate(err.obs,per_err=error*100)
err.rand<-mutate(err.rand,per_err=error*100)

#combine results
rf_results<-rbind.data.frame(err.obs,err.rand)

#results table 
(rf_results_tab<-rbind.data.frame(get_results("observed",err.obs$per_err),get_results("randomized",err.rand$per_err)))

#global results for Random forest classification of ACMA trees based on genotype. 
(res_plot<-ggplot(rf_results, aes(x=per_err, fill=group))+
  geom_density(alpha=0.7)+ 
  scale_fill_manual(values=c("grey80", "grey50"),labels=c("Observed Data", "Randomized Data")) +
  theme_classic()+ theme_bw()+
  scale_x_continuous(limits = c(0, 100))+
  #scale_y_continuous(limits = c(0, .2))+
  xlab("Classification Error (%)")+
  ylab("Kernal Density Estimate")+ #I think we can call this Frequency or ~Frequency, think about it.
  geom_vline(xintercept = mean(err.obs$per_err),color="blue")+
  geom_vline(xintercept = mean(err.rand$per_err),color="black")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.background = element_blank(),
        legend.key.size = unit(.4, "cm"),
        legend.text = element_text(size = 8),
        legend.key=element_rect(color="white")))

#now we will look at those results by class
err.class<-mutate(err.class,per_err=error*100)

#summary statistics for classification error by eco-region included in table 1.
(class_err_summary<-err.class[c(2,4)]%>%group_by(pop)%>%summarise_all(funs(mean,median,min,max)))

#a different color palette because when I use coord_flip the colors will no longer match.
backwards_colors<-c("#8dd3c7","#ffffb3","#bebada","#fb8072","#80b1d3","#fdb462","#b3de69","#fccde5","#d9d9d9","#bc80bd","#ccebc5","#ffed6f")

#boxplots showing classification error by eco-region (supplementary figure 6)

(class_err<-ggplot(aes(x=pop,y=per_err,fill=pop),color="black",data=err.class)+
  geom_boxplot(show.legend = FALSE,alpha=.6)+
  scale_fill_manual(values=backwards_colors)+
  #scale_x_discrete(labels=c("South America","Central America"))+ #use this line to change Eco-Region labels
  ylab(expression("Classification Error (%)"))+
  xlab("Eco-Region")+
  scale_y_continuous(limits = c(0, 100))+
  theme_bw()+
  theme(legend.position="none",
        panel.grid.minor = element_blank())+coord_flip())
